---
title: "Reproducible Research: Peer Assessment 1"
author: "Sean Corum"
date: "May 11, 2014"
output: html_document
---

## Loading and preprocessing the data
The R code below obtains the data.  The "...data [are from] from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

```{r}
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
download.file(url, "activity.zip", method = 'curl')
unzip('activity.zip')
time <- format(Sys.time(), '%a %b %d %X %Y')
data <- read.csv('activity.csv')
```

The data were obtained from `r url` on `r time`.

## What is mean total number of steps taken per day?

The total steps per day and the mean and median total steps per day can be computed
by the following R code:
```{r}
sumPerDay <- aggregate(steps ~ date, data = data, sum, na.action = na.omit)
meanSum <- mean(sumPerDay$steps, na.rm = TRUE)
meanSum
medianSum <- median(sumPerDay$steps, na.rm = TRUE)
medianSum
```

Ignoring missing values, the mean and median total steps per day are `r as.integer(meanSum)` and `r medianSum`, respectively.

The following is a histogram of the total steps per day:

```{r}
with(sumPerDay, hist(steps))
```

## What is the average daily activity pattern?

The following code makes a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
averagePerInterval <- aggregate(steps ~ interval, data = data, mean,
                                na.action = na.omit)
with(averagePerInterval, plot(interval, steps, type = 'l'))
```

The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is given by:

```{r}
averagePerInterval[max(averagePerInterval$steps) == averagePerInterval$steps,1]
```

## Imputing missing values

The total number of missing values in the dataset (i.e. the total number of rows with NAs) is given by:

```{r}
sum(!complete.cases(data))
```

The missing step values in the data set may be imputed by replacing them as with the median value of the same 5-minute interval across all days. A new dataset with the imputed values using this strategy may be constructed as follows:

```{r}
library(Hmisc)
newdata <- data
```

Including the imputed values, the total number of steps taken per day, as well as the mean and median total steps per day, may be computed as: 

```{r}
newSumPerDay <- aggregate(steps ~ date, data = newdata, sum)
newMeanSum <- mean(newSumPerDay$steps)
newMeanSum
newMedianSum <- median(newSumPerDay$steps)
newMedianSum
```

With imputed missing values, the mean and median total steps per day are `r as.integer(newMeanSum)` and `r newMedianSum`, respectively, which deviate from the original values by `r abs(newMeanSum-meanSum)/meanSum*100`% and `r abs(newMedianSum-medianSum)/medianSum*100`%, respectively.

The following is a histogram of the total steps per day the data with imputed missing values:

```{r}
with(newSumPerDay, hist(steps))
```

## Are there differences in activity patterns between weekdays and weekends?
A new factor variable may be created with two levels – “weekday” and “weekend”  - indicating whether a given date is a weekday or weekend day using the following R code:

```{r}
newdata$daytype <- weekdays(as.Date(as.character(newdata$date), '%Y-%m-%d'))
newdata$daytype[grep('Monday|Tuesday|Wednesday|Thursday|Friday',newdata$daytype)] <-   'weekday'
newdata$daytype[grep('Saturday|Sunday',newdata$daytype)] <- 'weekend'
newdata$daytype <- as.factor(newdata$daytype)
```

Then, the average across each interval for each type of day (weekday or weekend) may be found and plotted.
```{r}
averagePerIntervalAndDayType <- aggregate(steps ~ interval + daytype, data = newdata, mean, na.action = na.omit)
library(ggplot2)
qplot(interval, steps, data = averagePerIntervalAndDayType, facets = daytype ~., geom = 'line')
```

By inspection, the answer seems to be yes, there are differences in the average  pattern of steps per time on weekdays vs. weekends, with the most notible difference being a more variable number of steps being taken at different times throughout the day on the weekends. This would indicate that the activity associated with walking is less repetative on weekends than on weekdays.
